---
title: "NBA Sports Betting Analysis"
output: 
  flexdashboard::flex_dashboard:
    theme: 
      bootswatch: zephyr
    orientation: columns
    vertical_layout: fill
    source_code: embed
---

<style>
.chart-title {  /* chart_title  */
   font-size: 16px;
  }
body{ /* Normal  */
      font-size: 14px;
  }
</style>

```{r setup, include=FALSE}
library(flexdashboard)
library(pacman)

p_load(tidyverse, maps, viridis, ggplot2, plotly, htmltab, readr, leaflet, scales, plyr, reshape2, ggcorrplot, gridExtra)
```

Introduction
===

Column {data-width = 500} 
---

### A Brief History & Background Information

In 1931, Nevada becomes the first place in the United States where
gambling and betting on sports is legal. It remains the only place to do
so for decades. In 1975, the first true "sportsbook casino" is built in Las Vegas.
In 2018, the United States Supreme Court ruled that each state can freely establish their own sports gambling laws. Delaware stepped up first, joining Nevada, and by the end of the year, so did New Jersey, Rhode Island, Mississippi, West Virginia, New Mexico, Arkansas, and Pennsylvania. Also in 2018, popular online sportsbook DraftKings establishes itself in New Jersey. At the time, DraftKings was the first online sportsbook outside of Nevada. Currently, at time of writing, sports betting is legal with regulated industries in thirty-six states and Washington D.C. Two more states have anticipated launches in 2024 with five more reportedly close behind.
Popular online sportsbooks across the country include DraftKings, FanDuel, BetMGM, and Bet365.

### Why Analyze?

The idea behind this data set is to discover any trends between pre-game betting values, such as moneylines or spreads, and overall outcome of the game, like player points or winning team.

Column {data-width = 500} 
---

### Goals

The goal of this project is to explore the connection, if such exists, between a 
pre-game bet placements and the outcome of the game. Analyzing this data and finding a trend could produce better results when placing a bet on a basketball game.
and more success for the better. I believe this is an extremely topical discussion and exploration because sports betting is becoming increasingly legal and popular across the United States.

Questions this project aims to solve:

  - 1 How does the moneyline affect the final score/outcome of the game?
  
  - 2 How does the spread affect the final score/outcome of the game?
  
  - 3 Do the spread and moneyline correlate?
  
  - 4 Does the team that wins the tipoff win, more often?
  
  - 5 Does the team who makes the first basket win, more often?


Data Introduction
===

Column {.tabset data-width=550}
---

### Player Data

```{r}
playerdata <- read.csv("C:/Users/margo/OneDrive/Documents/MTH 369/nbabettinglines/nba_all_players_2021-22.csv")
teamdata <- read.csv("C:/Users/margo/OneDrive/Documents/MTH 369/nbabettinglines/teamdata.csv")

gamelogs2122 <- read.csv("C:/Users/margo/OneDrive/Documents/MTH 369/nbabettinglines/nba_gamelogs_2021-22.csv")

##odds1212 <- read.csv("C:/Users/margo/OneDrive/Documents/MTH 369/nbabettinglines/odds_fpts.csv")
##teamodds1212 <- read.csv("C:/Users/margo/OneDrive/Documents/MTH 369/nbabettinglines/odds_fpts_team.csv")

ftts1212 <- read.csv("C:/Users/margo/OneDrive/Documents/MTH 369/nbabettinglines/odds_ftts.csv")

outputfpts <- read.csv("C:/Users/margo/OneDrive/Documents/MTH 369/nbabettinglines/output_fpts.csv")
outputftts <- read.csv("C:/Users/margo/OneDrive/Documents/MTH 369/nbabettinglines/output_ftts.csv")

outputwintip <- read.csv("C:/Users/margo/OneDrive/Documents/MTH 369/nbabettinglines/output_win_tip.csv")

samplemoneyline <- read.csv("C:/Users/margo/OneDrive/Documents/MTH 369/nbabettinglines/nba_20221018_moneyline.csv")

samplespread <- read.csv("C:/Users/margo/OneDrive/Documents/MTH 369/nbabettinglines/nba_20221018_spread.csv")

moneylines <- read.csv("C:/Users/margo/OneDrive/Documents/MTH 369/nbabettinglines/allmoneylines.csv")
```

```{r}
playerdata <- playerdata[ , names(playerdata) %in% c("PLAYER_NAME", "TEAM_ABBREVIATION", "AGE", "PLAYER_HEIGHT_INCHES", "PLAYER_WEIGHT", "GP", "PTS", "REB", "AST", "NET_RATING", "OREB_PCT", "DREB_PCT", "USG_PCT", "TS_PCT", "AST_PCT", "SEASON_YEAR")]

teamdata <- teamdata[ , names(teamdata) %in% c("SEASON_YEAR", "TEAM_CITY", "TEAM_NAME", "TEAM_ABBREVIATION", "TEAM_CONFERENCE", "TEAM_DIVISION", "MIN_YEAR", "MAX_YEAR", "TEAM_FULL_NAME")]

samplemoneyline <- samplemoneyline[ , names(samplemoneyline) %in% c("game_id", "Date", "bet_type", "away_Team", "home_Team", "away_line", "home_line", "away_outcome", "home_outcome")]

samplespread <- samplespread[ , names(samplespread) %in% c("game_id", "Date", "bet_type", "away_Team", "home_Team", "away_line", "home_line")]

moneylines <- moneylines[ , names(moneylines) %in% c("game_id", "Date", "bet_type", "away_Team", "home_Team", "away_line", "home_line", "away_outcome", "home_outcome")]
```

```{r}
DT::datatable(playerdata, rownames = FALSE, 
              colnames = c("Player Name", "Team Abbreviation", "Age", "Height (in.)", "Weight", "Games Played", "Points", "Rebounds", "Assists", "Net Rating", "Offensive Rebound %", "Defensive Rebound %", "Usage %", "Total Scoring %", "Assist %", "Season"))
```

### Team Data
```{r}
DT::datatable(teamdata, rownames = FALSE, 
              colnames = c("Season", "City", "Name", "Abbreviation", "Conference", "Division", "Min Year", "Max Year", "Team Full Name"))
```

### Moneyline Example
```{r}
DT::datatable(samplemoneyline, rownames = FALSE, 
              colnames = c("Teams", "Date", "Bet Type", "Away Team", "Home Team", "Away Line", "Home Line", "Away Team Outcome", "Home Team Outcome"))
```

### Spread Example
```{r}
DT::datatable(samplespread, rownames = FALSE, 
              colnames = c("Teams", "Date", "Bet Type", "Away Team", "Home Team", "Away Line", "Home Line"))
```

### Variable Descriptions

**Player Data**

Player Name: Name of NBA player

Team Abbreviation: A unique three letter code that defines the team name

Age: Age of player, in years

Height: Height of player, in inches

Weight: Weight of player, in pounds

Games Played: How many games this player has played in over the course of the entire NBA season

Points: How many total points this player scored across the entire season

Rebounds: How many total rebounds this player accumulated over the season

Assists: How many total assists this player had over the entire season

Net Rating: 

Offensive Rebound %:

Defensive Rebound %:

Usage %:

Total Scoring %:

Assist %:

Season: Years that the basketball games were played


**Team Data**

Season: Years that the basketball games were played

City: The city in which the team plays in or near

Name: The mascot or team descriptor

Abbreviation: A unique three letter code that defines the team name

Conference: East or West, a location based descriptor that separates the teams into two groups

Division: blank, a more specific location based descriptor that divides each conference into

Min Year: The first year that the team participated in the NBA

Max Year: The most recent year that the team participated in the NBA


**Moneyline Example**

Teams: In a x_y format, each variable is a team abbreviation, describing who played the game

Date: In a YYYYMMDD format, the date in which the game took place

Bet Type: The type of bet being tracked (in this case, moneyline)

Away Team: The team in which is the visitor

Home Team: The team that is hosting the game in their city and arena

Away Line: 

Home Line:


**Spread Example**

Teams: In a x_y format, each variable is a team abbreviation, describing who played the game

Date: In a YYYYMMDD format, the date in which the game took place

Bet Type: The type of bet being tracked (in this case, spread)

Away Team: The team in which is the visitor

Home Team: The team that is hosting the game in their city and arena

Away Line: 

Home Line:


Sports Betting Basics
===

Column {.tabset data-width=550}
---

### Moneyline, Spread, etc.

explanations go here


Graphical Displays 1
===

Column {.tabset data-width=550}
---

### Moneylines

```{r, figures-side, fig.show="hold", out.width="50%"}
par(mfrow = c(2,2))

ggplot(moneylines, aes(x = away_line)) + geom_histogram(bins = 20, fill = "cornflowerblue") + labs(title = "Distribution of Away Moneylines", x = "Away Line", y = "Count")

ggplot(moneylines, aes(x = home_line)) + geom_histogram(bins = 25, fill = "cornflowerblue") + labs(title = "Distribution of Home Moneylines", x = "Home Line", y = "Count")

ggplot(moneylines, aes(x = as.factor(away_outcome), y = away_line)) + 
    geom_boxplot(fill = "royalblue1", alpha = 0.5) + labs(title = "Away Moneyline vs Outcome", x = "Win or Loss", y = "Moneyline")

ggplot(moneylines, aes(x = as.factor(home_outcome), y = home_line)) + 
    geom_boxplot(fill = "royalblue1", alpha = 0.5) + labs(title = "Home Moneyline vs Outcome", x = "Win or Loss", y = "Moneyline")
```

### Spreads



### Moneyline vs Spread?
find linear relationship or not?



### Win Tip



### Score First



Regression / Modeling
===

Column {.tabset data-width=550}
---

### tab title


Analysis / Displays 2
===

Column {.tabset data-width=550}
---

### tab title


Conclusion
===

Column {.tabset data-width=550}
---

### tab title


Author
===

Column {.tabset data-width=550}
---

### About the Author